Cartograms with Dynamic Features

### 2022-03-18 paper/section/introduction.tex
draft: review

Cartograms are representations of hybrid geographical and data space based on a value-by-area mapping technique that combines statistical and geographical information \cite{dent2009Cartography}. Various styles of cartograms have been proposed and implemented, covering applications such as urban planning \cite{harris2018Mapping, arranz-lopez2021Enduser}, natural hazard forecasting \cite{pappenberger2019Cartograms, park2020Flood}, conservation and environmental planning \cite{galluzzi2018Mapping, rocchini2019Cartogramming}, political and social demographics \cite{breitzman2018Using, alieva2021How}, and public health decision making \cite{gao2020Visualising, sack2021Visualizing}.

In this paper, we follow the categorization of cartograms by \citea{nusrat2016State}: contiguous, non-contiguous, rectangular, and Dorling. Among these four types, a trade-off between accuracy is made (See \tableref{table:accuracy}). For this project we focus on non-contiguous cartograms like Demers cartograms, because they facilitate statistical comparison between regions and they can make good use of screen space. Building on Demers cartograms \cite{ian2002Cartogram}, we introduce and develop novel dynamic topological features, such as rivers, aiming to improve the readability and geographical accuracy without sacrificing statistical accuracy. We implement a cartographic layout algorithm that includes these topological features into the layout of the nodes representing geographical regions. To minimize geographical errors and make efficient use of screen space, the algorithm also updates the position of rivers to accommodate the node layout. We then apply the algorithm a real world case-study using an Electronic Health Records (EHR) dataset to assess the accuracy of our algorithm.

>>>>

Cartograms are representations of hybrid geographical and abstract data based on a value-by-area mapping technique combining statistical and geographical information \cite{dent2009Cartography}. Various styles of cartograms have been proposed and implemented, covering applications such as urban planning \cite{harris2018Mapping, arranz-lopez2021Enduser}, natural hazard forecasting \cite{pappenberger2019Cartograms, park2020Flood}, conservation and environmental planning \cite{galluzzi2018Mapping, rocchini2019Cartogramming}, political and social demographics \cite{breitzman2018Using, alieva2021How}, and public health decision making \cite{gao2020Visualising, sack2021Visualizing}.

Among the four types of cartograms categorized by \citea{nusrat2016State} (contiguous, non-contiguous, rectangular, and Dorling), a trade-off between types of accuracy is made (See \tableref{table:accuracy}). For this project we focus on non-contiguous cartograms like Demers cartograms, because they facilitate statistical comparison between regions and they can make good use of screen space. Building on Demers cartograms \cite{ian2002Cartogram}, we introduce and develop novel dynamic topological features, such as rivers, aiming to improve the readability and geographical accuracy without sacrificing statistical accuracy. We implement a new cartographic layout algorithm that includes these topological features into the layout of the nodes representing geographical regions. To minimize geographical errors and make efficient use of screen space, the algorithm also updates the position of rivers to accommodate the node layout. One of the major challenges involves is how to develop a layout algorithm that handles different types of nodes. In other words, the layout algorithm is novel because it handles different types of nodes: rectangular representing regions and polylines representing rivers. We then apply the algorithm to a real world case-study using an Electronic Health Records (EHR) dataset to assess the accuracy of our algorithm. We also present a user study demonstrating its effectiveness.

### 2022-03-18 paper/section/main.tex
draft: review

{
\begin{figure}[tb!]
    \centering
    \includegraphics[width=\columnwidth]{figure/flowchart.png}
    \caption{A flowchart showing the abstraction of our layout algorithm.}
    \label{fig:flowchart}
\end{figure}
}

\algoref{alg:UpdateNodePosition} and \figref{fig:flowchart} provide an overview of our algorithm. \textbf{Initialization:} We first load and render the CCG geospatial boundaries. For each CCG we compute the centroid and represent it using a small unit size square node with $ \nodeSize $ as the initial size. \textbf{Node layout:} The basic algorithm is as follows: we first apply the Fast Node Overlap Removal (VPSC) algorithm \cite{dwyer2006fast} to remove overlaps. We chose VPSC over other node overlap removal algorithms since VPSC is able to provide spread minimization and node movement minimization while maintaining a good level of global shape preservation \cite{chen2020Node}. During overlap removal, we compute node trajectories (See \algoref{alg:check river intersection}) and translate nodes to their new position. Nodes that cross a river are translated back to their previous position. This procedure ends when 1) no node overlap is present; 2) no nodes cross a river. We then increase $ \nodeSize $ by one pixel and repeat the algorithm until the max node size $ \nodeSizeMax $ is reached. \new{The size growing process provides a stability in the layout and minimizes geographical errors.}

@@

    \textbf{Local variables:} \\
    $ \nodeListVPSC \gets $ a list of nodes representing CCGs after VPSC \\
    $ \nodeListCross \gets True $ if any node in the list crossed a river \\
    $ \nodeCross \gets True $ if the node crossed a river \\
    $ \nodeListEach \gets $ a node in $ \nodeList $ \\
    $ \nodeListEachPrevious \gets $ the previous position of $ \nodeListEach $\\
    $ \nodeListVPSCEach \gets $ a node in $ \nodeListVPSC $, which is the position of $ \nodeListEach $ after VPSC \\

@@

                \State $ \nodeListCross \gets False $
                \State $ \nodeListVPSC \gets $ RemoveOverlap ($ \nodeList $)

@@

We use rivers as topological boundaries and prevent nodes from crossing them. The resolution of rivers can be adjusted by the user, as shown in \figref{fig:river resolution}, the number of edges is reduced from 10,170 to 65. Enabling river simplification reduces the number of edge intersection tests that need to be performed. When a node's position changes, we test if the node's trajectory intersects any segment of a river. See \algoref{alg:check river intersection}. A bounding box intersection test between the edge defined by node translation and river edges can be performed to reduce the number of edge intersection detections required.

@@

As the VPSC always tries to produce an optimal node layout where node distribution and translation are minimized, a node's translation path might be repeatedly intersecting a river due to congestion, creating a stalemate situation, as shown in \figref{fig:stalemate}. If a node is translated between two positions for $ \stalemateMax $ iterations, a corridor is derived to break up the congestion. A corridor, $ c $, is a rectangle with a width of $ \CorridorWidth $ and a length of $ \CorridorLength $, formed by deriving two edges $ \EdgeParallelA $ and $ \EdgeParallelB $ such that $ \EdgeParallelA \parallel \EdgeParallelA \parallel \nodeLineNpP $ (See \figref{fig:corridor}C and D). All enclosed nodes are then translated by $ \nodeVectorPC $ to alleviate the congestion (See \figref{fig:corridor}D).

{
\begin{figure}[tb!]
    \centering
    \includegraphics[width=\columnwidth]{figure/stalemate.png}
    \caption{A stalemate situation is when a node's translation path $ \nodeVectorCP $ intersects a river for $ \stalemateMax $ times. A stalemate often occurs when the area is congested and the node is unable to move to a position that is not intersecting a river.}
    \label{fig:stalemate}
\end{figure}
}

{
\begin{figure}[tb!]
    \centering
    \includegraphics[width=\columnwidth]{figure/corridor.png}
    \caption{A stalemate occurs when a node's translation path $ \nodeVectorCP $ intersects a river for $ \stalemateMax $ times, as shown in A. To address the issue, we derive a corridor (orange rectangle in D) based on $ \node $ and $ \nodePrevious $. All nodes within the corridor are translated based on $ \nodeVectorPC $, such that $ \nodeVectorPC = \nodeVectorNNn = \nodeVectorNinNinn $. In order to show a clear illustration, we place nodes sparsely in this figure.}
    \label{fig:corridor}
\end{figure}
}

@@

    \textbf{Local variables:} \\
    $ \Corridor \gets $ the corridor \\
    $ \nodePrevious \gets $ the previous position of $ \node $ \\
    $ \PointP \gets $ the point extending $ \nodeVectorPC $ such that $ \nodeLineWidthNpP = \CorridorLength $\\
    $ \EdgeParallelA, \EdgeParallelB \gets $ the edges parallel to $ \nodeLineNpP $ \\
    $ corridor \gets $ a rectangle formed by $ \EdgeParallelA $ and $ \EdgeParallelB $ \\

>>>>

{
\begin{figure}[tb!]
    \centering
    \includegraphics[width=\columnwidth]{figure/flowchart.png}
    \caption{A flowchart showing our layout algorithm.}
    \label{fig:flowchart}
\end{figure}
}

\algoref{alg:UpdateNodePosition} and \figref{fig:flowchart} provide an overview of our algorithm. \textbf{Initialization:} We first load and render the CCG geospatial boundaries. For each CCG we compute the centroid and represent it using a small unit size square node with initial size, $ \nodeSize $. We start with small unit size squares because they do not overlap initially. \textbf{Node layout:} The basic algorithm is as follows: we first apply the Fast Node Overlap Removal (VPSC) algorithm \cite{dwyer2006fast} to remove overlaps. We chose VPSC over other node overlap removal algorithms because VPSC is able to provide spread minimization and node movement minimization while maintaining a good level of global shape preservation \cite{chen2020Node}. We gradually increase the node size by 1 unit at a time to ensure smooth transitions. An increase in size can cause the nodes to overlap. During overlap removal, we compute node trajectories (See \algoref{alg:check river intersection}) and translate nodes to their new position. Nodes that cross a river are translated back to their previous position. This procedure ends when 1) no node overlap is present; 2) no nodes cross a river. We then increase $ \nodeSize $ by one pixel and repeat the algorithm until the max node size $ \nodeSizeMax $ is reached. The gradual size increase process provides stability to the layout and helps minimize geographical error.

@@

    \textbf{Local variables:} \\
    $ \nodeListVPSC \gets $ a list of nodes representing CCGs after VPSC \\
    $ \nodeListCross \gets True $ if any node in the list crosses a river \\
    $ \nodeCross \gets True $ if node crosses a river \\
    % $ \nodeListEach \gets $ a node in $ \nodeList $ \\
    $ \nodeListEachPrevious \gets $ the previous position of $ \nodeListEach $\\
    $ \nodeListVPSCEach \gets $ a node in $ \nodeListVPSC $, which is the position of $ \nodeListEach $ after VPSC \\

@@

                \State \Comment{We should merge two lists into one}
                \State $ \nodeListCross \gets False $
                \State $ \nodeListVPSC \gets $ RemoveOverlap ($ \nodeList $)

@@

We use rivers as topological boundaries and prevent nodes from crossing them. The resolution of rivers can be adjusted by the user, as shown in \figref{fig:river resolution}, the number of edges is reduced from its original 10,170 to 65. We do not require a pixel-level river resolution as Demers cartograms do not represent geography at this level od detail, instead a resolution that matches the nodes is used. Enabling river simplification reduces the number of edge intersection tests that need to be performed. When a node's position changes, we test if the node's trajectory intersects any segment of a river. See \algoref{alg:check river intersection}. A bounding box intersection test between the edge defined by node translation and river edges can be performed to reduce the number of edge intersection detections required.

@@

As the VPSC always attempts to produce an optimal node layout where node distribution and translation are minimized, a node's translation path might be repeatedly intersecting a river due to congestion, creating a stalemate situation, as shown in \figref{fig:stalemate}. If a node is translated between two positions for $ \stalemateMax $ iterations, a corridor is constructed to break up the congestion. A corridor, $ c $, is a rectangle with a width of $ \CorridorWidth $ and a length of $ \CorridorLength $, formed by deriving two edges $ \EdgeParallelA $ and $ \EdgeParallelB $ such that $ \EdgeParallelA \parallel \EdgeParallelA \parallel \nodeLineNpP $ (See \figref{fig:corridor}C and D). All enclosed nodes are then translated by $ \nodeVectorNC $ to alleviate the congestion (See \figref{fig:corridor}D).

{
\begin{figure}[tb!]
    \centering
    \includegraphics[width=\columnwidth]{figure/stalemate.png}
    \caption{A stalemate situation is when a node's translation path $ \nodeVectorCN $ intersects a river for $ \stalemateMax $ times. A stalemate often occurs when the area is congested and the node is unable to move to a position that is not intersecting a river.}
    \label{fig:stalemate}
\end{figure}
}

{
\begin{figure}[tb!]
    \centering
    \includegraphics[width=\columnwidth]{figure/corridor.png}
    \caption{A stalemate occurs when a node's translation path $ \nodeVectorCN $ intersects a river for $ \stalemateMax $ times, as shown in A. To address the issue, we derive a corridor (orange rectangle in D) based on $ \node $ and $ \nodeVPSC $. All nodes within the corridor are translated based on $ \nodeVectorNC $, such that $ \nodeVectorNNn = \nodeVectorNinNinn $. In order to show a clear illustration, we place nodes sparsely in this figure.}
    \label{fig:corridor}
\end{figure}
}

@@

    \textbf{Local variables:} \\
    $ \Corridor \gets $ the corridor \\
    $ \nodePrevious \gets $ the previous position of $ \node $ (this here is confusing because $ \nodePrevious $ is actually $ \nodeVPSC $ in the previous iteration)\\
    $ \PointP \gets $ the point extending $ \nodeVectorNC $ such that $ \nodeLineWidthNtP = \CorridorLength $\\
    $ \EdgeParallelA, \EdgeParallelB \gets $ the edges parallel to $ \nodeLineNpP $ \\
    $ corridor \gets $ a rectangle formed by $ \EdgeParallelA $ and $ \EdgeParallelB $ \\

### 2022-03-18 paper/section/related.tex
draft: review

While we focus on rectangular and Dorling cartograms in this paper, we start with brief definitions of contiguous and non-contiguous cartograms for interested readers: Contiguous cartograms are topology-preserving that maintain connectivity with their adjacent neighbors but are also subject to distortion in shape. Non-contiguous cartograms sacrifice the topological connectivity with neighbors to enable expansion or reduction in size while maintaining their polygonal shape \cite{dent2009Cartography}.

\citea{nusrat2016State} define and summarize three major accuracy dimensions for cartograms: statistical, geographical, and topological. Each cartogram design may make accuracy trade-offs between dimensions. We provide an overview of these cartograms in Table \ref{table:accuracy}.

@@

\thumbnail{dorling}{
	Dorling cartograms are non-contiguous and do not preserve geography and topology. A Dorling cartogram is statistically accurate, regions are represented by circles and the data dimensions of interest are represented by the circle size \cite{dorling2011Area}. In a Demers cartogram, a variant of Dorling, squares are used instead to capture a certain level of topology \cite{cano2015Mosaic}. Dorling cartograms are unable to maintain topological accuracy as circles are often repositioned to remove overlap. Here we focus on Demers cartograms as we use squares to depict regions. This style of cartogram offers the advantages that the comparisons between regions are intuitive and screen space utilization is more efficient. Demers cartograms also have the advantage of preserving a higher level of topology. 
}

@@

Cartograms are a popular choice of representation in covering various topics by the media, with the goal of presenting an unbiased and complete picture of the news. The Washington Post uses cartograms to visualize the US overseas economic assistance, arm sales (Mosaic) \cite{bearak2016Everything}, the 2016 US Election (contiguous) \cite{gamio2016Election}, and the Brexit Referendum (Mosaic) \cite{taylor2016What}. National Geographic uses contiguous and Mosaic cartograms to analyze the 2016 US Election results \cite{miller2016Election}, the same topic is also covered by the Financial Times with a Dorling cartogram \cite{stabe2016Search}. \citea{cruz2017Adapted} adapts a Dorling cartogram with both contiguous and non-contiguous cartograms to represent the gender pay gap in Portugal. \citea{sandberg2018Cartogram} reports the 2018 US midterm Election with a Mosaic cartogram, the same approach is used to cover the 2020 US Election by the New York Times \cite{thelearningnetwork2020What} and Bloomberg \cite{mccartney20202020}.

>>>>

While we focus on rectangular and Dorling cartograms in this paper, we start with brief definitions of contiguous and non-contiguous cartograms: Contiguous cartograms are topology-preserving that maintain connectivity with their adjacent neighbors but are also subject to distortion in shape. Non-contiguous cartograms sacrifice the topological connectivity with neighbors to enable expansion or reduction in size while maintaining their polygonal shape \cite{dent2009Cartography}.

\citea{nusrat2016State} define and summarize three major accuracy dimensions for cartograms: statistical, geographical, and topological. Each cartogram design may make various types of accuracy trade-offs between dimensions. We provide an overview of these cartograms in Table \ref{table:accuracy}.

@@

\thumbnail{dorling}{
	Dorling cartograms are non-contiguous and do not preserve geography and topology. A Dorling cartogram is statistically accurate, regions are represented by circles and the data dimensions of interest are represented by the circle size \cite{dorling2011Area}. In a Demers cartogram, a variant of Dorling, squares are used instead to capture a certain level of topology \cite{cano2015Mosaic}. Dorling cartograms are unable to maintain topological accuracy as circles are often repositioned to remove overlap. Here we focus on Demers cartograms as we use squares to depict regions. This style of cartogram offers the advantages that the comparisons between regions are intuitive and screen space utilization is more efficient. This is important in our use case scenario involving EHRs. Demers cartograms also have the advantage of preserving a higher level of topology. 
}

@@

Cartograms are a popular choice of representation in covering various topics by the media, with the goal of presenting an unbiased and complete picture of the news. The Washington Post uses cartograms to visualize the US overseas economic assistance, in arm sales (Mosaic) \cite{bearak2016Everything}, the 2016 US Election (contiguous) \cite{gamio2016Election}, and the Brexit Referendum (Mosaic) \cite{taylor2016What}. National Geographic uses contiguous and Mosaic cartograms to analyze the 2016 US Election results \cite{miller2016Election}, the same topic is also covered by the Financial Times with a Dorling cartogram \cite{stabe2016Search}. \citea{cruz2017Adapted} adapts a Dorling cartogram with both contiguous and non-contiguous cartograms to represent the gender pay gap in Portugal. \citea{sandberg2018Cartogram} reports the 2018 US midterm Election with a Mosaic cartogram, the same approach is used to cover the 2020 US Election by the New York Times \cite{thelearningnetwork2020What} and Bloomberg \cite{mccartney20202020}.
